String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-09 2023-1-19/sections/cefa.tex
revise syntax in cefa

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

@@

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed.  We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
We define the semantics of a CEFA $\aut$, denoted by $\Lang(\aut)$, as $\{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$. Moreover, we define the \emph{output} of a CEFA $\aut$, denoted by $\cefaout(\aut)$, as $\{\myvec{v'} \mid  \exists w.\ \myvec{v'} \in \aut(w)\}$.

We would like to remark that the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion in the sense that the accepting condition does not distinguish final states. This technical choice is made so that determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

%
%and $\alpha(q_n)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}}$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings that are accepted by $\aut$.

@@

% Parikh's theorem \cite{parikh_theorem}\cite{parikh_anthony}\cite{parikh_compute} states that the Parikh image of a context-free language is semilinear so that it can be written as Presburger formula. Following the version for context-free grammar \cite{parikh_compute} and the version for regular expression \cite{parikh_for_nfa}, we define the Parikh image of language of an CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$  as follows:
% \begin{table}[h]
%   \begin{tabular}{l l r}
%     $\psi(\aut)\equiv$& $\bigwedge\limits_{q\in Q}(\text{$f_q$ if $q\in F$ otherwise 0}) + \sum\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}} = $ & \\
%     & \quad \ \ (1 if $q = q_I$ otherwise 0) + $\sum\limits_{(q',a,q,\myvec{v})\in\delta} t_{q',a,q, \myvec{v}} $ & \\ 
%     & $\bigwedge\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}}\geq 0$ & (consistent formula)\\
%     & & \\
%     & $\bigwedge\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}}>0\rightarrow z_{q'} > 0$ & \\
%     & $\bigwedge\limits_{q\in F} z_q > 0\rightarrow f_q = 1$ & \\ 
%     & $\bigwedge\limits_{q\in Q} z_q > 0\rightarrow \bigvee\limits_{(q,a,q',\myvec{v})\in\delta} z_q = z_{q'} + 1\wedge t_{q,a,q',\myvec{v}}\geq 0 \wedge z_{q'} > 0$& (connected formula)\\
%     & & \\
%     &$\bigwedge\limits_{r_i\in R} r_i = \sum\limits_{(q,a,q',\myvec{v})\in \delta} t_{q,a,q',\myvec{v}}*\myvec{v}[i]$ & (register value formula)\\
%     & & \\
%     &$\bigwedge\limits_{q\in F} f_q > 0 \rightarrow \theta(q)$ & (accepting condition) 
%   \end{tabular}
% \end{table}
% $z_q$ is the distance of state $q$ from $q_f\in F$ in a spanning tree. $t_{q,a,q',\myvec{v}}$ represents the using times of transition $(q,a,q',\myvec{v})$ in the accepting run. $f_q$ stands for whether an accepting state $q\in F$ is the final state of a run. The Parikh image of $\aut$ is a Presburger formula $\psi(\aut)$, which is consistent and connected. The register value formula ensures the register values are correctly updated regarding the accepting run. 
% \begin{example} \label{exapmle:parikh}
%   Suppose that we have a CEFA $\aut$ shown in Fig. \ref{subfig:aut_x}. Using $t_1, t_2, t_3$ to record the times of transitions in the accepting run of $\aut$, the Parikh image $\psi(\aut)$ is the quantifier-free Presburger formula $\psi_{cons}\wedge \psi_{conn} \wedge \psi_{val} \wedge \psi_{aut}$ where:
%   \begin{itemize}
%     \item $\psi_{cons} \equiv 1 = t_1 \wedge t_1 + t_3 = t_2 \wedge t_2 = t_3 + 1 \bigwedge\limits_{i\in[1,3]} t_i\geq 0$ is the consistent formula.
%     \item $\psi_{conn} \equiv \top $ is the connectivity formula because all transitions in $\aut$ are connected.
%     \item $\psi_{val} \equiv r_1=t_1 + t_3\wedge r_2 = t_1+t_2+t_3$ is the formula deciding the value of $r_1$ and $r_2$ after a run.
%     \item $\psi_{aut} \equiv r_2 = i\wedge 1 \leq r_1 \leq 100$ is the accepting condition of $\aut$.
%   \end{itemize}
%   $\psi(\aut)$ can be simplified to $t_3\geq 0 \wedge t_1=1\wedge r_1 = t_3 + 1\wedge r_2 = 2*t_3+2\wedge r_2=i\wedge 1\leq r_1\leq 100$. A possible solution is $t_1 =1, t_3 = 0, r_1=1, r_2=2$ and the corresponding accepted string $ab$. It is obvious that $r_1$ is the repeat times of $ab$ and $r_2$ is the length of $ab$.
% \end{example}
%\end{document}

>>>>

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add\denghang{adds} write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

@@

The semantics of CEFA is define\denghang{defined} as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note here \denghang{that} we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
We define the semantics of a CEFA $\aut$, denoted by $\Lang(\aut)$, as $\{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$. Moreover, we define the \emph{output} of a CEFA $\aut$, denoted by $\cefaout(\aut)$, as $\{\myvec{v'} \mid  \exists w.\ \myvec{v'} \in \aut(w)\}$.

We would like to remark that the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion in the sense that the accepting condition does not distinguish final states. This technical choice is made so that \denghang{the} determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

%
%and $\alpha(q_n)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}}$. Note here \denghang{that} we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings that are accepted by $\aut$.

@@

% Parikh's theorem \cite{parikh_theorem}\cite{parikh_anthony}\cite{parikh_compute} states that the Parikh image of a context-free language is semilinear so that it can be written as Presburger formula. Following the version for context-free grammar \cite{parikh_compute} and the version for regular expression \cite{parikh_for_nfa}, we define the Parikh image of language of an CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$  as follows:
% \begin{table}[h]
%   \begin{tabular}{l l r}
%     $\psi(\aut)\equiv$& $\bigwedge\limits_{q\in Q}(\text{$f_q$ if $q\in F$ otherwise 0}) + \sum\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}} = $ & \\
%     & \quad \ \ (1 if $q = q_I$ otherwise 0) + $\sum\limits_{(q',a,q,\myvec{v})\in\delta} t_{q',a,q, \myvec{v}} $ & \\ 
%     & $\bigwedge\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}}\geq 0$ & (consistent formula)\\
%     & & \\
%     & $\bigwedge\limits_{(q,a,q',\myvec{v})\in\delta} t_{q,a,q',\myvec{v}}>0\rightarrow z_{q'} > 0$ & \\
%     & $\bigwedge\limits_{q\in F} z_q > 0\rightarrow f_q = 1$ & \\ 
%     & $\bigwedge\limits_{q\in Q} z_q > 0\rightarrow \bigvee\limits_{(q,a,q',\myvec{v})\in\delta} z_q = z_{q'} + 1\wedge t_{q,a,q',\myvec{v}}\geq 0 \wedge z_{q'} > 0$& (connected formula)\\
%     & & \\
%     &$\bigwedge\limits_{r_i\in R} r_i = \sum\limits_{(q,a,q',\myvec{v})\in \delta} t_{q,a,q',\myvec{v}}*\myvec{v}[i]$ & (register value formula)\\
%     & & \\
%     &$\bigwedge\limits_{q\in F} f_q > 0 \rightarrow \theta(q)$ & (accepting condition) 
%   \end{tabular}
% \end{table}
% $z_q$ is the distance of state $q$ from $q_f\in F$ in a spanning tree. $t_{q,a,q',\myvec{v}}$ represents the using times of transition $(q,a,q',\myvec{v})$ in the accepting run. $f_q$ stands for whether an accepting state $q\in F$ is the final state of a run. The Parikh image of $\aut$ is a Presburger formula $\psi(\aut)$, which is consistent and connected. The register value formula ensures \denghang{that} the register values are correctly updated regarding the accepting run. 
% \begin{example} \label{exapmle:parikh}
%   Suppose that we have a CEFA $\aut$ shown in Fig. \ref{subfig:aut_x}. Using $t_1, t_2, t_3$ to record the times of transitions in the accepting run of $\aut$, the Parikh image $\psi(\aut)$ is the quantifier-free Presburger formula $\psi_{cons}\wedge \psi_{conn} \wedge \psi_{val} \wedge \psi_{aut}$ where:
%   \begin{itemize}
%     \item $\psi_{cons} \equiv 1 = t_1 \wedge t_1 + t_3 = t_2 \wedge t_2 = t_3 + 1 \bigwedge\limits_{i\in[1,3]} t_i\geq 0$ is the consistent formula.
%     \item $\psi_{conn} \equiv \top $ is the connectivity formula because all transitions in $\aut$ are connected.
%     \item $\psi_{val} \equiv r_1=t_1 + t_3\wedge r_2 = t_1+t_2+t_3$ is the formula deciding the value of $r_1$ and $r_2$ after a run.
%     \item $\psi_{aut} \equiv r_2 = i\wedge 1 \leq r_1 \leq 100$ is the accepting condition of $\aut$.
%   \end{itemize}
%   $\psi(\aut)$ can be simplified to $t_3\geq 0 \wedge t_1=1\wedge r_1 = t_3 + 1\wedge r_2 = 2*t_3+2\wedge r_2=i\wedge 1\leq r_1\leq 100$. A possible solution is $t_1 =1, t_3 = 0, r_1=1, r_2=2$ and the corresponding accepted string $ab$. It is obvious that $r_1$ is the repeat times of $ab$ and $r_2$ is the length of $ab$.
% \end{example}
%\end{document}
